---
title: "NOAA Storm Data Analysis"
output: 
  html_document: 
    keep_md: true
---

##Synopsis
In this report we show which weather events have a greatest impact to injury and health as well as the greatest economic impact overall for the United States. In order to investigate this hypothesis we have gathered the data from [U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database.](http://www.ncdc.noaa.gov/stormevents/) The events in the database start in the year 1950 and end in November 2011. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. Based on the details captured on the [NOAA website](http://www.ncdc.noaa.gov/stormevents/details.jsp) only data from 1996 onwards will be studied. This is to stop bias from being introduced into the data. From this data taking the top 10 events, we found Tornado's are major danger to public health as it results in more fatalities and Injuries. Flood events on the other hand have had major impact on economy of the United States.


#Data Processing

###Get and load the data
This section gets and loads the required data. If the data already has already been downloaded, the data is not downloaded again.
```{r Loading and Preprocessing the data, cache=TRUE} 
# set the url path for the download 
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

#set the working directory
setwd("~/R/RepData_PeerAssessment2")

# set up some working variables. These can be changed depending on your requirements
workingDir = getwd()
dataDir = "data"
fileName = "data/StormData.csv.bz2"


# create a data directory if one does not exist directory
if (!file.exists(dataDir)){
  dir.create(file.path(workingDir, dataDir))
}

# This setting is used to for https connections. Windows machine no curl installed
setInternet2(use = TRUE) #This needs to be set to support https. Could not use curl as not installed

if (!file.exists(fileName)){
  
  # download into the data directory
  download.file(fileUrl, fileName, method = "auto")
}


# load the csv in data frame
rawStormData <- read.csv(fileName, as.is=TRUE)
```

###Check the dataset

Check the first few rows of the dataset
```{r}
head(rawStormData)
```

Check the column names of the activity dataset

```{r}
names(rawStormData)
```

Check the structure of the activity dataset 
```{r}
str(rawStormData)
```


###Transformations and Preprocessing the dataset

Make all the column names lower case. This makes it easier to read and program.
```{r}
names(rawStormData) <- tolower(make.names(names(rawStormData)))
```

Change the begin date to be a proper date
```{r}
rawStormData$bgn_date <- as.Date(rawStormData$bgn_date, format="%m/%d/%Y") #Australian date format
```

In the next section we select only the data that is required. Based on the details captured on the [NOAA website](http://www.ncdc.noaa.gov/stormevents/details.jsp) only data from 1996 onwards will be studied. This is to stop bias from being introduced into the data.
```{r}
stormData <- subset(rawStormData, bgn_date >= as.Date("01-01-1996"), select=c(bgn_date, countyname, state, evtype, fatalities, injuries, propdmg, propdmgexp, cropdmg, cropdmgexp))

head(stormData)

unique(stormData[, "evtype"])

```


###Combining the event types

From the above unique event type values we can see that there are a fairly large number. It will be best to combine similar events for our data analysis. Convert all the values of the event type (evtype) column to upper case values and then combine similar event types.
```{r}
stormData[,c("evtype")] <- toupper(stormData[,c("evtype")])


stormData[stormData$evtype == "AVALANCE", c("evtype")] <- "AVALANCHE"

stormData[grep("BLIZZARD*", stormData$evtype), c("evtype")] <- "BLIZZARD"

stormData[grep("HAIL*", stormData$evtype), c("evtype")] <- "HAIL"

stormData[grep("HEAVY RAIN*", stormData$evtype), c("evtype")] <- "HEAVY RAIN"

stormData[grep("WATERSPOUT*", stormData$evtype), c("evtype")] <- "WATERSPOUT"

stormData[grep("HURRICANE*", stormData$evtype), c("evtype")] <- "HURRICANE"

stormData[grep("THUNDERSTORM*|TUNDERSTORM WIND*|TSTM WIND*|THUDERSTORM WINDS*", stormData$evtype), 
    c("evtype")] <- "THUNDERSTORM WIND"

stormData[grep("THUNDEERSTORM WINDS*", stormData$evtype), c("evtype")] <- "THUNDERSTORM WIND"

stormData[grep("THUNDERESTORM WINDS*", stormData$evtype), c("evtype")] <- "THUNDERSTORM WIND"

stormData[grep("THUNDERTORM WINDS*", stormData$evtype), c("evtype")] <- "THUNDERSTORM WIND"

stormData[grep("THUNERSTORM WINDS*", stormData$evtype), c("evtype")] <- "THUNDERSTORM WIND"

stormData[grep("THUNDERSTROM WIND*", stormData$evtype), c("evtype")] <- "THUNDERSTORM WIND"

stormData[grep("THUNDERSTROM WIND*", stormData$evtype), c("evtype")] <- "THUNDERSTORM WIND"

stormData[grep("TSTMW*", stormData$evtype), c("evtype")] <- "THUNDERSTORM WIND"

stormData[grep("TORNADO*", stormData$evtype), c("evtype")] <- "TORNADO"

stormData[grep("TORNDAO*", stormData$evtype), c("evtype")] <- "TORNADO"

stormData[grep("RIP CURRENT*", stormData$evtype), c("evtype")] <- "RIP CURRENT"

stormData[grep("STRONG WIND*", stormData$evtype), c("evtype")] <- "STRONG WIND"

stormData[grep("LIGHTNING*", stormData$evtype), c("evtype")] <- "LIGHTNING"

stormData[grep("LIGHTING*|LIGNTNING*", stormData$evtype), c("evtype")] <- "LIGHTNING"

stormData[grep("FLASH FLOOD*", stormData$evtype), c("evtype")] <- "FLASH FLOOD"

stormData[grep("WINTER WEATHER*", stormData$evtype), c("evtype")] <- "WINTER WEATHER"

stormData[grep("WINTER STORM*", stormData$evtype), c("evtype")] <- "WINTER STORM"

stormData[grep("TROPICAL STORM*", stormData$evtype), c("evtype")] <- "TROPICAL STORM"

stormData[grep("HEAVY SNOW*", stormData$evtype), c("evtype")] <- "HEAVY SNOW"

stormData[grep("HEAVY RAIN*|HVY RAIN*", stormData$evtype), c("evtype")] <- "HEAVY RAIN"

stormData[grep("FLOOD/FLASH*|FLOOD FLASH*", stormData$evtype), c("evtype")] <- "FLASH FLOOD"

stormData[grep("FLOODING|FLOOD/RIVER FLOOD|FLOODS|FLOOD/RAIN/WINDS", stormData$evtype), c("evtype")] <- "FLOOD"

stormData[grep("WILDFIRES*|WILD FIRES*|WILDFIRE*|WILD/FOREST*", stormData$evtype), c("evtype")] <- "WILDFIRE"

stormData[grep("HURRICANE*|TYPHOON*", stormData$evtype), c("evtype")] <- "HURRICANE (TYPHOON)"


stormData$totalInjuriesFatalities <- stormData$injuries + stormData$fatalities
head(sort(tapply(stormData$totalInjuriesFatalities, stormData$evtype, sum), decreasing=TRUE ), n = 20)

```

###Get the economical data in the correct formats
Lets look the unique values from property damage exponent and crop property damage exponents.  

```{r}
unique(stormData[, "propdmgexp"])

unique(stormData[, "cropdmgexp"])
```
First we need to change the exponent values from charaters to numericals before calculating the total amount of damage. I convert the property damage and crop damage data into comparable numerical form, based on the meaning of units described in the code book (National Climatic Data Center's [record layout document](http://ire.org/media/uploads/files/datalibrary/samplefiles/Storm%20Events/layout08.doc), which is referenced on the [Investigative Reporers & Editors web site](http://ire.org/nicar/database-library/databases/storm-events/).
```{r}
#Correct the property damage exponent
stormData$propdmgexp = as.character(stormData$propdmgexp)
stormData$propdmgexp[toupper(stormData$propdmgexp) == "B"] = "9"
stormData$propdmgexp[toupper(stormData$propdmgexp) == "M"] = "6"
stormData$propdmgexp[toupper(stormData$propdmgexp) == "K"] = "3"
stormData$propdmgexp[toupper(stormData$propdmgexp) == "H"] = "2"
stormData$cropdmgexp[toupper(stormData$cropdmgexp) == ""] = "0"
stormData$cropdmgexp[toupper(stormData$cropdmgexp) == "-"] = "0"
stormData$cropdmgexp[toupper(stormData$cropdmgexp) == "+"] = "0"
stormData$cropdmgexp[toupper(stormData$cropdmgexp) == "?"] = "0"
stormData$propdmgexp = as.numeric(stormData$propdmgexp)
stormData$propdmgexp[is.na(stormData$propdmgexp)] = 0
stormData$totalPropertyDamage = stormData$propdmg * 10^stormData$propdmgexp
summary(stormData$totalPropertyDamage)

#correct the crop damage exponent
stormData$cropdmgexp = as.character(stormData$cropdmgexp)
stormData$cropdmgexp[toupper(stormData$cropdmgexp) == "B"] = "9"
stormData$cropdmgexp[toupper(stormData$cropdmgexp) == "M"] = "6"
stormData$cropdmgexp[toupper(stormData$cropdmgexp) == "K"] = "3"
stormData$cropdmgexp[toupper(stormData$cropdmgexp) == "H"] = "2"
stormData$cropdmgexp[toupper(stormData$cropdmgexp) == ""] = "0"
stormData$cropdmgexp[toupper(stormData$cropdmgexp) == "?"] = "0"
stormData$cropdmgexp[is.na(stormData$cropdmgexp)] = 0
stormData$cropdmgexp = as.numeric(stormData$cropdmgexp)
stormData$totalCropDamage = stormData$cropdmg * 10^stormData$cropdmgexp
summary(stormData$totalCropDamage)

#calculate the total damage
stormData$totalDamage <- stormData$totalPropertyDamage + stormData$totalCropDamage
```




#Results
This section provides insights in the results that were discussed in the Synopsis

###Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

The following code generates a barplot for the top 10 weather events that cause injuries and fatalities by the number.The requirement is determine which events have the greatest consequences with respect to population health, keeping this in mind we are only going to select the top 10 events.
```{r}
#Get the total fatalities and injuries for each evetn type
injuryFatalityData <- tapply(stormData$totalInjuriesFatalities, stormData$evtype, sum)
#create a data frame from the above calculation
injuryFatalityData <- data.frame(evtype = names(injuryFatalityData), totalInjuriesFatalities = as.numeric(injuryFatalityData))
#order the data frame by the total number of injuries/fatalities from largest to smallest
injuryFatalityData <- injuryFatalityData[order(injuryFatalityData$totalInjuriesFatalities, decreasing = TRUE),]
#select on the top 10
injuryFatalityData <- injuryFatalityData[1:10,]
# get the top event that has caused the most injuries and fatalities
top1FatalityEvent <- injuryFatalityData[1, "evtype"]
#set the margins
par(mar = c(13, 7, 2, 2), las = 3)
#plot the 
barplot(injuryFatalityData$totalInjuriesFatalities, names.arg = injuryFatalityData$evtype, col = injuryFatalityData$evtype, ylab = "Total # of Injuries/Fatalities", 
    main = "Top 10 Events That Cause Injuries or Fatalities")

title(xlab = "Event Type", line = 11)
```


The graph above shows the total number of fatalities and injuries by event. Only the top 10 events are shown in the graph. Based on the above plot the `r top1FatalityEvent` event causes the most injuries and fatalities

##Across the United States, which types of events have the greatest economic consequences?
The following code generates a barplot for the top 25 weather events that cause injuries and fatalities by the number. The requirement is determine which events have the greatest economic consequences, keeping this in mind we are only going to select the top 10 events.
```{r}
#Get the total damage for each event type
totalDamageData <- tapply(stormData$totalDamage, stormData$evtype, sum)
#create a data frame from the above calculation
totalDamageData <- data.frame(evtype = names(totalDamageData), totalDamage = as.numeric(totalDamageData)/1000000000) #set as billion
#order the data frame by the total damage from largest to smallest
totalDamageData <- totalDamageData[order(totalDamageData$totalDamage, decreasing = TRUE),]
#select on the top 10
totalDamageData <- totalDamageData[1:10,]
# get the top event that has caused the most damage
top1DamageEvent <- totalDamageData[1, "evtype"]
#set the margins
par(mar = c(13, 7, 2, 2), las = 3)
#create a bar plot
barplot(totalDamageData$totalDamage, names.arg = totalDamageData$evtype, col = totalDamageData$evtype, ylab = "Total Damage In Billions", 
    main = "Top 25 Events Taht Cause The Most Economical Damage")

title(xlab = "Event Type", line = 11)
```

The graph above shows the total damage in billions of dollars by event. Only the top 10 events are shown in the graph. From the Graph above, the `r top1DamageEvent` event causes the economical impact in the United States

#Session and Environment Information
```{r}
sessionInfo()
```